Cognate beginnings to bilingual lexical acquisition

### 2023-07-24 manuscript/manuscript.qmd
Add new paragraph to discussion

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together.

Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome].

@@

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological for similar approaches].

Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor.

@@

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients. Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. *p*(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero. In the right column, we show a graphical representation of the median and 95% HDI of each coefficient, with a reference dotted line indicating the location of zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

@@

>>>>

We collected data for `r format(nrow(n_items), big.mark = ",")` words. We restricted the analyses to responses to nouns [`r nrow(n_items)-nrow(n_items_nouns)` items corresponding to other grammatical classes were excluded, see @fourtassi2020growth for a similar approach]. We then excluded items with missing lexical frequency scores (*n* = `r nrow(n_items_nouns)-nrow(n_items_freq)`), items that included more than one lemma (e.g., *mono/mico* [monkey], *n* = `r nrow(n_items_freq)-nrow(n_items_lemma)`), multi-word items or phrases (e.g., *barrita de cereales* [cereal bar], *n* = `r nrow(n_items_lemma)-nrow(n_items_multi)`). Finally, we removed items without a translation in the other language (*n* = `r nrow(n_items_multi)-nrow(n_items_duplicated)`). This resulted in a final list of `r sum(n_items_final)` items, corresponding to `r n_items_final["Catalan"]` Catalan words and their `r n_items_final["Spanish"]` Spanish translations (`r n_te` translation equivalents). After collecting participants' responses, the final dataset consisted of `r format(n_obs, big.mark = ",")` observations, each corresponding to a single response of one participant to one item. Each translation equivalent received a median of `r round(sum_obs$median)` responses (*Min* = `r round(sum_obs$min)`, *Max* = `r round(sum_obs$max)`) from participants, both languages pooled together. Data processing and visualisation was done in R [@rcoreteam2013language, version `r r_version`] using the Tidyverse family of packages [@wickham2019welcome].

@@

We defined *Cognateness* as the phonological similarity between each word-form and its translation. For each translation equivalent, we used the `stringdist` [@vanderloo2014stringdist] R package to calculate the Levenshtein distance between the Catalan and the Spanish phonological transcriptions of the word-forms. The Levenshtein distance measures the number of editions (insertions, deletions, or substitutions) that one string of phonemes/characters must go through to become identical to the other [@levenshtein1966binary]. We divided the Levenshtein distance of each translation equivalent by the length of the longest word-form to correct for word length (longer strings are likely to show a larger number of mismatches). Finally, we subtracted the result from one so that it could be interpreted in terms of phonological similarity, instead of phonological distance. This led to a distance metric that ranged from zero to one, where zero indicates that both word-forms are completely different (e.g., /\textipa{"taw.l5}/--/\textipa{"me.sa}/, *table*), and one indicates that the two word-forms are identical (e.g., /\textipa{"mar}/--/\textipa{"mar}/, *sea*) [see @floccia2018introduction; @heeringa2003norwegian; @schepens2012distributions; @fourtassi2020growth; @laing2022phonological; for similar approaches]. Predictors were standardised before entering the model by subtracting the mean of the predictor from each value and dividing the result by the standard deviation of the predictor.

@@

```{r tbl-coefs}
#| label: tbl-coefs
#| tbl-cap: "Posterior distribution of regression coefficients. Median: median of the posterior distribution in the probability scale. 95\\% HDI: 95\\% highest density interval of the distribution. *p*(ROPE): overlap between the 95\\% HDI and the ROPE, indicating the posterior probability that the true value of the coefficient is equivalent to zero."
#| echo: false
#| message: false
#| warning: false
#| out-width: 80%
# summarise posterior draws
intercept_label <- "Intercepts (at 22 months)"

@@

The asymmetric facilitation of cognateness on word acquisition reported in the present study parallels previous findings in adults. For  instance, unbalanced (or low-proficiency) bilingual adults benefit more strongly from forward (dominance to non-dominant) cross-language phonological priming during word processing than from backward phonological priming (non-dominant to dominant) [e.g., @grainger1998masked; @voga2007cognate; @de1991lexical; @dimitropoulou2011masked]. On the contrary, more balanced (or high-proficiency bilinguals) show an equivalent priming facilitation in both directions [@basnight2007differences; @dunabeitia2009masked]. Similarly, the infants in the present study benefited from the cognate status of words in their less dominant language, but not in their more dominant language. It is possible that in both cases, the underlying cause behind the asymmetric effect of parallel activation lies in lexical representations in the non-dominant language receiving stronger activation from their translations in the dominant language than vice versa.

This dominance-mediated difference in connection strength is contemplated by previous models of bilingual lexical access, such as the Revised Hierarchical Model [RHM; @kroll1994category], but the specific implementation of the source of such difference is still unclear.  A more recent proposal, the Ontogenic Model of L2 Lexical Representation [@bordag2022ontogenesis; @cook2016fuzzy], points to a 'fuzzier'---or more poorly phonological specification---of lexical representations in the non-dominant second language, compared to those in the dominant language. This sub-optimal specification may be due to a mismatch between the phonological inventory of both languages [@cutler2006asymmetric; @sebastian2005influence; @sebastian2006first] or, more generally, an imprecise encoding of the features of the representation in any domain (e.g., form-meaning mapping, orthographical, phonological). Although this account seems convenient to explain cross-language asymmetries in language processing, it should be taken into account that most (if not all) of the proposals aforementioned are aimed at explaining results in adults, and not in children. Language dominance and proficiency are frequently defined in both populations using different dimensions. For instance, the low proficiency bilinguals in many of the aforementioned studies acquired their second language years after their toddlerhood. This might have changed substantially the developmental trajectory of language acquisition they followed, compared to the bilinguals in the present study, who were exposed to both languages from early. Moreover, our findings can in principle be explained in the absence of a poorer specification of representations in the non-dominant language. As suggested in the introduction, it is possible that simply the fact that infants are exposed more frequently to words in the dominant language leads to words in the non-dominant language receiving additional parallel activation, compared to words in the dominant language. This explanation does not necessarily compete with words in the dominant language activating more strongly their translations than vice versa, and further research is needed in order to clarify this issue.

### 2023-07-24 manuscript/manuscript.tex
Simplify dependencies
Add new paragraph to discussion

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

We collected data for 1,590 words. We restricted the analyses to
responses to nouns (628 items corresponding to other grammatical classes
were excluded, see \protect\hyperlink{ref-fourtassi2020growth}{Fourtassi
et al., 2020} for a similar approach). We then excluded items with
missing lexical frequency scores (\emph{n} = 269), items that included
more than one lemma (e.g., \emph{mono/mico} {[}monkey{]}, \emph{n} =
48), multi-word items or phrases (e.g., \emph{barrita de cereales}
{[}cereal bar{]}, \emph{n} = 9). Finally, we removed items without a
translation in the other language (\emph{n} = 32). This resulted in a
final list of 604 items, corresponding to 302 Catalan words and their
302 Spanish translations (302 translation equivalents). After collecting
participants' responses, the final dataset consisted of 139,326
observations, each corresponding to a single response of one participant
to one item. Each translation equivalent received a median of 234
responses (\emph{Min} = 108, \emph{Max} = 880) from participants, both
languages pooled together.

Data processing and visualisation was done in R
(\protect\hyperlink{ref-rcoreteam2013language}{R Core Team, 2013},
version 4.2.2) using the Tidyverse family of packages
(\protect\hyperlink{ref-wickham2019welcome}{Wickham et al., 2019}).

@@

We defined \emph{Cognateness} as the phonological similarity between
each word-form and its translation. For each translation equivalent, we
used the \texttt{stringdist}
(\protect\hyperlink{ref-vanderloo2014stringdist}{Loo, 2014}) R package
to calculate the Levenshtein distance between the Catalan and the
Spanish phonological transcriptions of the word-forms. The Levenshtein
distance measures the number of editions (insertions, deletions, or
substitutions) that one string of phonemes/characters must go through to
become identical to the other
(\protect\hyperlink{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided the Levenshtein distance of each translation equivalent by the
length of the longest word-form to correct for word length (longer
strings are likely to show a larger number of mismatches). Finally, we
subtracted the result from one so that it could be interpreted in terms
of phonological similarity, instead of phonological distance. This led
to a distance metric that ranged from zero to one, where zero indicates
that both word-forms are completely different (e.g.,
/\textipa{"taw.l5}/--/\textipa{"me.sa}/, \emph{table}), and one
indicates that the two word-forms are identical (e.g.,
/\textipa{"mar}/--/\textipa{"mar}/, \emph{sea}) (see
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-fourtassi2020growth}{Fourtassi et al., 2020};
\protect\hyperlink{ref-heeringa2003norwegian}{Heeringa \& Gooskens,
2003}; \protect\hyperlink{ref-laing2022phonological}{Laing, 2022} for
similar approaches;
\protect\hyperlink{ref-schepens2012distributions}{Schepens et al.,
2012}).

Predictors were standardised before entering the model by subtracting
the mean of the predictor from each value and dividing the result by the
standard deviation of the predictor.

@@

\hypertarget{tbl-coefs}{}
\begin{table}
\caption{\label{tbl-coefs}Posterior distribution of regression coefficients. Median: median of the
posterior distribution in the probability scale. 95\% HDI: 95\% highest
density interval of the distribution. \emph{p}(ROPE): overlap between
the 95\% HDI and the ROPE, indicating the posterior probability that the
true value of the coefficient is equivalent to zero. In the right
column, we show a graphical representation of the median and 95\% HDI of
each coefficient, with a reference dotted line indicating the location
of zero. }\tabularnewline

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

We collected data for 1,590 words. We restricted the analyses to
responses to nouns (628 items corresponding to other grammatical classes
were excluded, see \protect\hyperlink{ref-fourtassi2020growth}{Fourtassi
et al., 2020} for a similar approach). We then excluded items with
missing lexical frequency scores (\emph{n} = 269), items that included
more than one lemma (e.g., \emph{mono/mico} {[}monkey{]}, \emph{n} =
48), multi-word items or phrases (e.g., \emph{barrita de cereales}
{[}cereal bar{]}, \emph{n} = 9). Finally, we removed items without a
translation in the other language (\emph{n} = 32). This resulted in a
final list of 604 items, corresponding to 302 Catalan words and their
302 Spanish translations (302 translation equivalents). After collecting
participants' responses, the final dataset consisted of 139,326
observations, each corresponding to a single response of one participant
to one item. Each translation equivalent received a median of 234
responses (\emph{Min} = 108, \emph{Max} = 880) from participants, both
languages pooled together. Data processing and visualisation was done in
R (\protect\hyperlink{ref-rcoreteam2013language}{R Core Team, 2013},
version 4.2.2) using the Tidyverse family of packages
(\protect\hyperlink{ref-wickham2019welcome}{Wickham et al., 2019}).

@@

We defined \emph{Cognateness} as the phonological similarity between
each word-form and its translation. For each translation equivalent, we
used the \texttt{stringdist}
(\protect\hyperlink{ref-vanderloo2014stringdist}{Loo, 2014}) R package
to calculate the Levenshtein distance between the Catalan and the
Spanish phonological transcriptions of the word-forms. The Levenshtein
distance measures the number of editions (insertions, deletions, or
substitutions) that one string of phonemes/characters must go through to
become identical to the other
(\protect\hyperlink{ref-levenshtein1966binary}{Levenshtein, 1966}). We
divided the Levenshtein distance of each translation equivalent by the
length of the longest word-form to correct for word length (longer
strings are likely to show a larger number of mismatches). Finally, we
subtracted the result from one so that it could be interpreted in terms
of phonological similarity, instead of phonological distance. This led
to a distance metric that ranged from zero to one, where zero indicates
that both word-forms are completely different (e.g.,
/\textipa{"taw.l5}/--/\textipa{"me.sa}/, \emph{table}), and one
indicates that the two word-forms are identical (e.g.,
/\textipa{"mar}/--/\textipa{"mar}/, \emph{sea}) {[}see Floccia et al.
(\protect\hyperlink{ref-floccia2018introduction}{2018}); Heeringa \&
Gooskens (\protect\hyperlink{ref-heeringa2003norwegian}{2003}); Schepens
et al. (\protect\hyperlink{ref-schepens2012distributions}{2012});
Fourtassi et al. (\protect\hyperlink{ref-fourtassi2020growth}{2020});
Laing (\protect\hyperlink{ref-laing2022phonological}{2022}); for similar
approaches{]}. Predictors were standardised before entering the model by
subtracting the mean of the predictor from each value and dividing the
result by the standard deviation of the predictor.

@@

\hypertarget{tbl-coefs}{}
\begin{table}
\caption{\label{tbl-coefs}Posterior distribution of regression coefficients. Median: median of the
posterior distribution in the probability scale. 95\% HDI: 95\% highest
density interval of the distribution. \emph{p}(ROPE): overlap between
the 95\% HDI and the ROPE, indicating the posterior probability that the
true value of the coefficient is equivalent to zero. }\tabularnewline

@@

The asymmetric facilitation of cognateness on word acquisition reported
in the present study parallels previous findings in adults. For
instance, unbalanced (or low-proficiency) bilingual adults benefit more
strongly from forward (dominance to non-dominant) cross-language
phonological priming during word processing than from backward
phonological priming (non-dominant to dominant)
(\protect\hyperlink{ref-de1991lexical}{De Groot \& Nas, 1991};
\protect\hyperlink{ref-dimitropoulou2011masked}{Dimitropoulou et al.,
2011}; e.g., \protect\hyperlink{ref-grainger1998masked}{Grainger, 1998};
\protect\hyperlink{ref-voga2007cognate}{Voga \& Grainger, 2007}). On the
contrary, more balanced (or high-proficiency bilinguals) show an
equivalent priming facilitation in both directions
(\protect\hyperlink{ref-basnight2007differences}{Basnight-Brown \&
Altarriba, 2007};
\protect\hyperlink{ref-dunabeitia2009masked}{Duñabeitia et al., 2009}).
Similarly, the infants in the present study benefited from the cognate
status of words in their less dominant language, but not in their more
dominant language. It is possible that in both cases, the underlying
cause behind the asymmetric effect of parallel activation lies in
lexical representations in the non-dominant language receiving stronger
activation from their translations in the dominant language than vice
versa.

This dominance-mediated difference in connection strength is
contemplated by previous models of bilingual lexical access, such as the
Revised Hierarchical Model {[}RHM; Kroll \& Stewart
(\protect\hyperlink{ref-kroll1994category}{1994}){]}, but the specific
implementation of the source of such difference is still unclear. A more
recent proposal, the Ontogenic Model of L2 Lexical Representation
(\protect\hyperlink{ref-bordag2022ontogenesis}{Bordag et al., 2022};
\protect\hyperlink{ref-cook2016fuzzy}{Cook et al., 2016}), points to a
`fuzzier'---or more poorly phonological specification---of lexical
representations in the non-dominant second language, compared to those
in the dominant language. This sub-optimal specification may be due to a
mismatch between the phonological inventory of both languages
(\protect\hyperlink{ref-cutler2006asymmetric}{Cutler et al., 2006};
\protect\hyperlink{ref-sebastian2006first}{Sebastian-Gallés et al.,
2006}; \protect\hyperlink{ref-sebastian2005influence}{Sebastián-Gallés
et al., 2005}) or, more generally, an imprecise encoding of the features
of the representation in any domain (e.g., form-meaning mapping,
orthographical, phonological). Although this account seems convenient to
explain cross-language asymmetries in language processing, it should be
taken into account that most (if not all) of the proposals
aforementioned are aimed at explaining results in adults, and not in
children. Language dominance and proficiency are frequently defined in
both populations using different dimensions. For instance, the low
proficiency bilinguals in many of the aforementioned studies acquired
their second language years after their toddlerhood. This might have
changed substantially the developmental trajectory of language
acquisition they followed, compared to the bilinguals in the present
study, who were exposed to both languages from early. Moreover, our
findings can in principle be explained in the absence of a poorer
specification of representations in the non-dominant language. As
suggested in the introduction, it is possible that simply the fact that
infants are exposed more frequently to words in the dominant language
leads to words in the non-dominant language receiving additional
parallel activation, compared to words in the dominant language. This
explanation does not necessarily compete with words in the dominant
language activating more strongly their translations than vice versa,
and further research is needed in order to clarify this issue.

@@

\leavevmode\vadjust pre{\hypertarget{ref-basnight2007differences}{}}%
Basnight-Brown, D. M., \& Altarriba, J. (2007). Differences in semantic
and translation priming across languages: The role of language direction
and language dominance. \emph{Memory \& Cognition}, \emph{35}, 953--965.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bordag2022ontogenesis}{}}%
Bordag, D., Gor, K., \& Opitz, A. (2022). Ontogenesis model of the L2
lexical representation. \emph{Bilingualism: Language and Cognition},
\emph{25}(2), 185--201.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cook2016fuzzy}{}}%
Cook, S. V., Pandža, N. B., Lancaster, A. K., \& Gor, K. (2016). Fuzzy
nonnative phonolexical representations lead to fuzzy form-to-meaning
mappings. \emph{Frontiers in Psychology}, \emph{7}, 1345.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cutler2006asymmetric}{}}%
Cutler, A., Weber, A., \& Otake, T. (2006). Asymmetric mapping from
phonetic to lexical representations in second-language listening.
\emph{Journal of Phonetics}, \emph{34}(2), 269--284.

\leavevmode\vadjust pre{\hypertarget{ref-de1991lexical}{}}%
De Groot, A. M., \& Nas, G. L. (1991). Lexical representation of
cognates and noncognates in compound bilinguals. \emph{Journal of Memory
and Language}, \emph{30}(1), 90--123.

\leavevmode\vadjust pre{\hypertarget{ref-dimitropoulou2011masked}{}}%
Dimitropoulou, M., Duñabeitia, J. A., \& Carreiras, M. (2011). Masked
translation priming effects with low proficient bilinguals. \emph{Memory
\& Cognition}, \emph{39}, 260--275.

\leavevmode\vadjust pre{\hypertarget{ref-dunabeitia2009masked}{}}%
Duñabeitia, J. A., Perea, M., \& Carreiras, M. (2009). Masked
translation priming effects with highly proficient simultaneous
bilinguals. \emph{Experimental Psychology}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-grainger1998masked}{}}%
Grainger, J. (1998). Masked priming by translation equivalents in
proficient bilinguals. \emph{Language and Cognitive Processes},
\emph{13}(6), 601--623.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kroll1994category}{}}%
Kroll, J. F., \& Stewart, E. (1994). Category interference in
translation and picture naming: Evidence for asymmetric connection
between bilingual memory representations. \emph{Journal of Memory and
Language}, \emph{33}(2), 149--174.
\url{https://doi.org/10.1006/jmla.1994.1008}

@@

\leavevmode\vadjust pre{\hypertarget{ref-sebastian2006first}{}}%
Sebastian-Gallés, N., Rodrı́guez-Fornells, A., Diego-Balaguer, R. de, \&
Dı́az, B. (2006). First-and second-language phonological representations
in the mental lexicon. \emph{Journal of Cognitive Neuroscience},
\emph{18}(8), 1277--1291.

\leavevmode\vadjust pre{\hypertarget{ref-sebastian2005influence}{}}%
Sebastián-Gallés, N., Echeverrı́a, S., \& Bosch, L. (2005). The influence
of initial exposure on lexical representation: Comparing early and
simultaneous bilinguals. \emph{Journal of Memory and Language},
\emph{52}(2), 240--255.

@@

\leavevmode\vadjust pre{\hypertarget{ref-voga2007cognate}{}}%
Voga, M., \& Grainger, J. (2007). Cognate status and cross-script
translation priming. \emph{Memory \& Cognition}, \emph{35}, 938--952.
